Should I AirBNB My Property?

### 2016-12-11 abbReturns_Script.Rmd
REORG:  Preliminary commands

---
title: "Should I AirBNB My Property?"
author: "Andy Krause and Gideon Aschwanden"
date: "1 November 2016"
output:
  html_notebook: default
  pdf_document: default
---

@@

This document is intended the literate programming of our analysis for the 'Should I AirBNB My Property?' research project.  Below we document the complete data provenance, taking our data from the raw source data through to the final results. Data prepartion, modeling and visualization will be done in R (version 3.3.1). As a very brief overview, we will be comparing the financial returns of short-term (AirBNB-type) and long-term (traditional leases) leasing of residential properties in Melbourne, Australia.  We perform this comparison under a hypothetical condition where-in an investor purchases a property on Septemeber 1, 2015 and is looking to determine whether too lease out the property on a traditional long-term lease (12 months) or as an AirBNB property.  We examine the returns for the following 12 months.  Sensitivity to time frames, start dates, locations, property types, etc. are considered in this analysis as well.

Data on AirBNB (short-term) rentals has been purchased from www.airdna.co, a data provider specializing in AirBNB data collection and analysis. The data on long-term leases was provided by Australia Property Monitors (APM), a Fairfax Group company.

Throughout this document we will refer to the short-term data as **'short-term'** or **'AirBNB'** while the long-term data will be referred to as **'long-term'**, **'rental'** or **'lease'**.  Short-term nightly fees will be referenced as **'rates'** and long-term as **'rent'**.  We will use **'prices'** to collectively refer to **rates** and **rents**. Note that within the Melbourne market rentals are quoted in weekly rents (but are charged to tenants at a monthly rent equivalent to the 1/12 of 52 weeks of rent).

@@

We begin with a number of preliminary commands that include setting the global options, loading necessary packages (libraries), setting the path to the raw data sources and reading in custom functions used later in the analysis.

```{r prelim_cmds, message=FALSE, warning=FALSE, comment=FALSE, echo=TRUE, results='hide'}

 ## Set global options

  options(xtable.comment = FALSE)

 ## Load libraries

@@

 ## Set path location

  data.path <- 'c:/dropbox/research/airBNB/data/'

 ## Load sources (custom function files)

  source("c:/code/dataviztools/ggplottools.R")
  source("c:/code/datamgmttools/dataMungetools.R")
  source("c:/code/research/AirBNBMelbourne/analysis_Functions.R")
  source("c:/code/research/AirBNBMelbourne/dataPrep_Functions.R")

@@

@@

  rent.data <- read.csv(paste0(data.path, '/prepared/ltpropdata.csv'), header=T)

@@

  rent.keep.flds <- c('id.key', 'geographicalid', 'eventid', 'addressid', 'activityid',
                      'flatnumber', 'streetnumber', 'streetname', 'streettype', 
                      'suburb', 'postcode', 'property_latitude',
                      'property_longitude', 'street_centroid_latitude',
                      'street_centroid_longitude', 'eventdate', 'eventprice',
                      'propertytype', 'areasize', 'bedrooms', 'baths', 'parking',
                      'hasstudy', 'hascourtyard', 'hasbalcony', 'hasairconditioning',
                      'hasgarage', 'dom')

  rent.data <- rent.data[ , rent.keep.flds]

@@

  # Limit to extent of data
  sub.table <- table(rent.dataf$suburb)
  sub.table <- sub.table[sub.table != 0]

@@

  # Convert to point shapefile
  rent.prop.shp <- SpatialPointsDataFrame(cbind(rent.dataf$longitude,
                                                rent.dataf$latitude),
                                         rent.dataf,
                                         proj4string=CRS("+init=epsg:4283"))

  # Add suburb name 
  spJoin <- over(rent.prop.shp, suburbs.shp)
  rent.dataf$suburb <- as.character(spJoin$name_2006)
  rent.dataf$suburb[is.na(rent.dataf$suburb)] <- 'missing'

  # Add suburb designation
  rent.dataf$sub.class <- as.character(spJoin$suburbdesi)
  rent.dataf$sub.class[is.na(rent.dataf$sub.class)] <- 'missing'

  # Add SA1 designation
  spJoin <- over(rent.prop.shp, sa1s.shp)
  rent.dataf$sa1 <- spJoin$sa1_main11

  ## Add Beach Designation
  spJoin <- over(rent.prop.shp, beach.buffer)
  rent.dataf$beach <- ifelse(is.na(spJoin), 0 ,1)

  # Remove those missing a suburb classification (out of area)
  rent.dataf <- rent.dataf[rent.dataf$sub.class != 'missing', ]
  rent.dataf <- rent.dataf[rent.dataf$sa1 != 'missing', ]

@@

  # Make map
  rent.map <- ggplot() +
    geom_polygon(data=subs, aes(x=long, y=lat, group=group), color='gray50') +
    geom_point(data=rent.dataf, aes(x=longitude, y=latitude, color=sub.mrkt), size=.1) +
    scale_color_manual(values=sm.col) +
    scale_x_continuous(limits=c(min(subs$long), max(subs$long))) +
    scale_y_continuous(limits=c(min(subs$lat), max(subs$lat))) +
    xlab('') + ylab('')+
    theme(legend.position = 'bottom',
          legend.title = element_blank()) +
    guides(colour = guide_legend(override.aes = list(size=3))) + 
    ggtitle('Long Term Rental Location\nBy Submarket')

@@

  # Make map
  renttype.map <- ggplot() +
    geom_bar(data=rent.dataf, aes(x=sub.mrkt, fill=sub.mrkt)) +
    scale_fill_manual(values=sm.col) +
    facet_wrap(~type) +
    ylab('# of Properties\n') + xlab('')+
    theme(legend.position = 'bottom',
          legend.title = element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle('Long Term Property Type\nBy Submarket')

@@

  rentbed <- ggplot(rent.dataf, aes(x=bedrooms)) + 
    geom_bar(fill=abb.col[3], color=abb.col[3]) +
    facet_wrap(~type) +
    scale_x_continuous(labels=1:10, breaks=1:10)+
    xlab('\nBedrooms\n') +
    ylab('') +
    ggtitle('Bedrooms in Long Term Rentals')

@@

  abba <- which(abb.dataf$bedrooms > 3 & abb.dataf$type =='Apartment')
  renta <- which(rent.dataf$bedrooms > 3 & rent.dataf$type =='Apartment')
  if(length(abba)>0) abb.dataf <- abb.dataf[-abba, ]
  if(length(renta)>0) rent.dataf <- rent.dataf[-renta, ]

@@

  ltbath <- ggplot(rent.dataf, aes(x=baths)) + 
    geom_bar(fill=abb.col[3], color=abb.col[3]) +
    facet_wrap(~type) +
    scale_x_continuous(labels=1:8, breaks=1:8)+
    xlab('\nBathrooms\n') +
    ylab('') + 
    ggtitle('Bathrooms in Long Term Rentals')

@@

  ltb <- ggplot(rent.dataf, aes(x=bedbath)) + 
    geom_bar(fill=abb.col[3], color=abb.col[3]) +
    xlab('\nBed/Bath Combo\n') +
    ylab('Number of Properties\n') +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ggtitle('Long Term Rentals') + 
    facet_wrap(~type)

@@

  # Rent Plot
  rent.plot <- ggplot(rent.dataf, aes(x=event.price)) + 
    geom_density(fill=abb.col[3],alpha=.8) +
    ylab('') +
    xlab('Weekly Rent') + 
    ggtitle('Long Term Rentals') + 
    scale_x_continuous(limits=c(0, 1500)) +
    theme(axis.title.y=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank())

@@

  rent.dataf <- rent.dataf[rent.dataf$event.price >= 200 & 
                         rent.dataf$event.price <= 1000, ]

@@

  listing.dataf <- listing.dataf[listing.dataf$activity.id %in% rent.dataf$activity.id,]

@@

>>>>

---
title: "Should I AirBNB My Property?"
author: "Andy Krause and Gideon Aschwanden"
date: "1 November 2016"
output:
  pdf_document: default
  html_notebook: default
---

@@

This document is the literate programming of our analysis for the 'Should I AirBNB My Property?' research project.  Below we document the complete data provenance, taking our data from the raw source data through to the final results. Data prepartion, modeling and visualization will be done in R (version 3.3.1). As a very brief overview, we will be comparing the financial returns of short-term (AirBNB-type) and long-term (traditional leases) leasing of residential properties in Melbourne, Australia.  We perform this comparison under a hypothetical condition where-in an investor purchases a property on Septemeber 1, 2015 and is looking to determine whether to lease out the property on a traditional long-term lease (12 months) or nightly as an AirBNB property.  We examine the returns for the following 12 months.  Sensitivity to time frames, start dates, locations, property types, etc. are considered in this analysis as well.

Data on AirBNB (short-term) rentals have been purchased from www.airdna.co, a data provider specializing in AirBNB data collection and analysis. The data on long-term leases was provided by Australia Property Monitors (APM), a Fairfax Group company.

Throughout this document we will refer to the short-term data as **'short-term'** or **'AirBNB'** while the long-term data will be referred to as **'long-term'**, **'rental'** or **'lease'**.  Short-term nightly fees will be referenced as **'rates'** and long-term as **'rent'**.  We will use **'prices'** to collectively refer to **rates** and **rents**. Note that within the Melbourne long term rentals are quoted in weekly rents but are charged to tenants at a monthly rent equivalent to 1/12 of 52 weeks of rent.

@@

We begin with a number of preliminary commands.  First we load all of the necessary R libraries for this analysis.

```{r load_libraries, message=FALSE, warning=FALSE, comment=FALSE, echo=TRUE, results='hide'}

@@

We then set the paths to the data and the code, conditional on which of the authors is running the analysis.

```{r set_paths, message=FALSE, warning=FALSE, comment=FALSE, echo=TRUE, results='hide'}

  # Get computer names

  comp.name <- Sys.info()['nodename']

  # Assign path based on computer name

  if(comp.name == '7020D-121777-W' | comp.name == 'DESKTOP-1D7JO4J'){

    data.path <- 'c:/dropbox/research/airBNB/data/'
    code.path <- 'c:/code/'

  } else {

    data.path <- 'alternate data path'
    code.path <- 'alternate data path'

  }

Next, we source the necessary custom code files for the analysis.

```{r source_files, message=FALSE, warning=FALSE, comment=FALSE, echo=TRUE, results='hide'}

  source(paste0(code.path, "dataviztools/ggplottools.R"))
  source(paste0(code.path, "datamgmttools/dataMungetools.R"))
  source(paste0(code.path, "research/AirBNBMelbourne/analysis_Functions.R"))
  source(paste0(code.path, "research/AirBNBMelbourne/dataPrep_Functions.R"))

Finally, we set the constants used in this study:

1. **Exchange rate**: The Airbnb rates are provided in US dollars.  We convert these to Australian dollars based on the average exchange rate over the Sep 1 2015 to Aug 30 2015 period, or about 1.32 Australian dollars to 1 US dollar.

```{r set_constants, message=FALSE, warning=FALSE, comment=FALSE, echo=TRUE, results='hide'}

  exch.rate <- 1.32

@@

Airbnb data in US dollars, so we must set an exchange rate.  Average of that period was 1.32 Australian dollars to US dollars.

```{r exch_rate, message=FALSE, warning=FALSE, comment=FALSE, echo=FALSE}

  exch.rate <- 1.32

@@

  ltr.data <- read.csv(paste0(data.path, '/prepared/ltpropdata.csv'), header=T)

@@

  ltr.keep.flds <- c('id.key', 'geographicalid', 'eventid', 'addressid', 'activityid',
                      'flatnumber', 'streetnumber', 'streetname', 'streettype', 
                      'suburb', 'postcode', 'property_latitude',
                      'property_longitude', 'street_centroid_latitude',
                      'street_centroid_longitude', 'eventdate', 'eventprice',
                      'propertytype', 'areasize', 'bedrooms', 'baths', 'parking',
                      'hasstudy', 'hascourtyard', 'hasbalcony', 'hasairconditioning',
                      'hasgarage', 'dom')

  ltr.data <- ltr.data[ ,ltr.keep.flds]

@@

  # Limit to extent of data
  sub.table <- table(ltr.dataf$suburb)
  sub.table <- sub.table[sub.table != 0]

@@

  # Convert to point shapefile
  rent.prop.shp <- SpatialPointsDataFrame(cbind(ltr.dataf$longitude,
                                                ltr.dataf$latitude),
                                         ltr.dataf,
                                         proj4string=CRS("+init=epsg:4283"))

  # Add suburb name 
  spJoin <- over(rent.prop.shp, suburbs.shp)
  ltr.dataf$suburb <- as.character(spJoin$name_2006)
  ltr.dataf$suburb[is.na(ltr.dataf$suburb)] <- 'missing'

  # Add suburb designation
  ltr.dataf$sub.class <- as.character(spJoin$suburbdesi)
  ltr.dataf$sub.class[is.na(ltr.dataf$sub.class)] <- 'missing'

  # Add SA1 designation
  spJoin <- over(rent.prop.shp, sa1s.shp)
  ltr.dataf$sa1 <- spJoin$sa1_main11

  ## Add Beach Designation
  spJoin <- over(rent.prop.shp, beach.buffer)
  ltr.dataf$beach <- ifelse(is.na(spJoin), 0 ,1)

  # Remove those missing a suburb classification (out of area)
  ltr.dataf <- ltr.dataf[ltr.dataf$sub.class != 'missing', ]
  ltr.dataf <- ltr.dataf[ltr.dataf$sa1 != 'missing', ]

@@

  # Make map
  rent.map <- ggplot() +
    geom_polygon(data=subs, aes(x=long, y=lat, group=group), color='gray50') +
    geom_point(data=ltr.dataf, aes(x=longitude, y=latitude, color=sub.mrkt), size=.1) +
    scale_color_manual(values=sm.col) +
    scale_x_continuous(limits=c(min(subs$long), max(subs$long))) +
    scale_y_continuous(limits=c(min(subs$lat), max(subs$lat))) +
    xlab('') + ylab('')+
    theme(legend.position = 'bottom',
          legend.title = element_blank()) +
    guides(colour = guide_legend(override.aes = list(size=3))) + 
    ggtitle('Long Term Rental Location\nBy Submarket')

@@

  # Make map
  renttype.map <- ggplot() +
    geom_bar(data=ltr.dataf, aes(x=sub.mrkt, fill=sub.mrkt)) +
    scale_fill_manual(values=sm.col) +
    facet_wrap(~type) +
    ylab('# of Properties\n') + xlab('')+
    theme(legend.position = 'bottom',
          legend.title = element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    ggtitle('Long Term Property Type\nBy Submarket')

@@

  rentbed <- ggplot(ltr.dataf, aes(x=bedrooms)) + 
    geom_bar(fill=abb.col[3], color=abb.col[3]) +
    facet_wrap(~type) +
    scale_x_continuous(labels=1:10, breaks=1:10)+
    xlab('\nBedrooms\n') +
    ylab('') +
    ggtitle('Bedrooms in Long Term Rentals')

@@

  abba <- which(abb.dataf$bedrooms > 3 & abb.dataf$type =='Apartment')
  renta <- which(ltr.dataf$bedrooms > 3 & ltr.dataf$type =='Apartment')
  if(length(abba)>0) abb.dataf <- abb.dataf[-abba, ]
  if(length(renta)>0) ltr.dataf <- ltr.dataf[-renta, ]

@@

  ltbath <- ggplot(ltr.dataf, aes(x=baths)) + 
    geom_bar(fill=abb.col[3], color=abb.col[3]) +
    facet_wrap(~type) +
    scale_x_continuous(labels=1:8, breaks=1:8)+
    xlab('\nBathrooms\n') +
    ylab('') + 
    ggtitle('Bathrooms in Long Term Rentals')

@@

  ltb <- ggplot(ltr.dataf, aes(x=bedbath)) + 
    geom_bar(fill=abb.col[3], color=abb.col[3]) +
    xlab('\nBed/Bath Combo\n') +
    ylab('Number of Properties\n') +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ggtitle('Long Term Rentals') + 
    facet_wrap(~type)

@@

  # Rent Plot
  rent.plot <- ggplot(ltr.dataf, aes(x=event.price)) + 
    geom_density(fill=abb.col[3],alpha=.8) +
    ylab('') +
    xlab('Weekly Rent') + 
    ggtitle('Long Term Rentals') + 
    scale_x_continuous(limits=c(0, 1500)) +
    theme(axis.title.y=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank())

@@

  ltr.dataf <- ltr.dataf[ltr.dataf$event.price >= 200 & 
                         ltr.dataf$event.price <= 1000, ]

@@

  listing.dataf <- listing.dataf[listing.dataf$activity.id %in% ltr.dataf$activity.id,]

@@

## Calculate Observed Revenues

```{r abb_revs, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

 ## Calculate actual revenue for study period for AirBnb

  # Determine # of booking for each airbnb
  abb.dataf$bookings <- (abb.dataf$total.days - 
                          abb.dataf$blocked.days - 
                           abb.dataf$available.days)

  # Determine property specific occupancy rate
  abb.dataf$occ.rate <- (abb.dataf$bookings / 
                           (abb.dataf$total.days - abb.dataf$blocked.days))

  # Set the blocked rate
  abb.dataf$blocked.rate <- abb.dataf$blocked.days / abb.dataf$total.days

  # Set the extrapolation parameter for those on the site for less than whole period
  abb.dataf$extr.par <- 366 / abb.dataf$total.days

  # Get all of the reservations
  resv.daily <- daily.dataf[daily.dataf$status == 'R' &
                             daily.dataf$booked.date != 'imputed' &
                              daily.dataf$price < 600, ]

  # Calculate the property specific median nightly rate
  med.daily <- tapply2DF(resv.daily$price, resv.daily$property.id, median)

  # Add median rate to the property  
  abb.dataf$med.rate <- med.daily$Var[match(abb.dataf$property.id, med.daily$ID)]

  # Calculate the total revenue  
  abb.dataf$revenue <- (abb.dataf$med.rate * exch.rate * 
                          abb.dataf$extr.par * abb.dataf$bookings)

  # Remove those with no revenue
  abb.dataf <- abb.dataf[!is.na(abb.dataf$revenue), ]
```

```{r ltr_revs, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

 ## Calculate ltral revenues

  # Calculate the actual revenue
  ltr.dataf$revenue <- ltr.dataf$event.price * (52 - ltr.dataf$dom / 7)

  # Remove those with very low revenues (due to very long DOMs)
  ltr.dataf <- ltr.dataf[ltr.dataf$revenue > 5000, ]

Data transformation steps

```{r data_trans3, message=FALSE, warning=FALSE, comment=FALSE, cache=TRUE, echo=TRUE}

  # Re-order factors
  abb.dataf$bedbath <- factor(abb.dataf$bedbath, 
                              levels=c('2..2', '1..1', '2..1', '3..1', '3..2', '4..2'))
  ltr.dataf$bedbath <- factor(ltr.dataf$bedbath, 
                               levels=c('2..2', '1..1', '2..1', '3..1', '3..2', '4..2'))

  # Add product as a variable
  abb.dataf$product <- paste0(substr(abb.dataf$type, 1, 1), abb.dataf$bedbath)
  ltr.dataf$product <- paste0(substr(ltr.dataf$type, 1, 1), ltr.dataf$bedbath)

  # Convert suburbs from factor to character
  abb.dataf$suburb <- as.character(abb.dataf$suburb)
  ltr.dataf$suburb <- as.character(ltr.dataf$suburb)

## Save for future costs
